Word Learning as Network Growth: A Cross-linguistic Analysis

### 2019-02-05 journal/ms.tex
modif journal

What factors shape vocabulary learning over the course of early
childhood? To investigate this question, scientists have adopted
multiple research strategies, from conducting controlled laboratory
experiments (e.g. Markman, 1990) to analyzing dense corpora capturing
language learning in context (e.g., B. C. Roy, Frank, DeCamp, Miller, \&
Roy, 2015). One strategy consists in documenting the timeline of words'
acquisition, and studying the properties that make words easy or hard to
learn. For example, within a lexical category, words that are more
frequent in child-directed speech are acquired earlier (J. C. Goodman,
Dale, \& Li, 2008). Other factors include word length and the mean
length of utterances in which the word occurs (e.g., Braginsky,
Yurovsky, Marchman, \& Frank, 2016; Swingley \& Humphrey, 2018).

Besides these word-level properties, the structure of the lexicon (that
is, how words relate to one another) also predicts the Age of
Acquisition (AoA) of words. The lexical structure can be characterized
in terms of a network where each node represents a word in the
vocabulary, and each link between two nodes represents a relationship
between the corresponding pair of words (e.g., Collins \& Loftus, 1975).
Previous studies have investigated early vocabulary structure by
constructing networks using a variety of word-word relations including
shared semantic features (McRae, Cree, Seidenberg, \& McNorgan, 2005),
target-cue relationships in free association norms (Nelson, McEvoy, \&
Schreiber, 1998), co-occurrence in child directed speech (MacWhinney,
2014), and phonological relatedness (Vitevitch, 2008). These studies
have generally found that children tend to produce words that have
higher neighborhood density (i.e., high connectivity in the network)
earlier, both at the phonological and the semantic level (Carlson,
Sonderegger, \& Bane, 2014; Hills, Maouene, Riordan, \& Smith, 2010;
Hills, Maouene, Maouene, Sheya, \& Smith, 2009; Stella, Beckage, \&
Brede, 2017; Storkel, 2009).

@@

These two proposals represent two divergent ideas about the role of
lexical networks in acquisition. On the INT proposal, learning is driven
by known words with high connectivity to other known words (Figure 1,
left). Thus, the network structure is a causal factor in early word
learning, that is, children rely on the organization of their past
knowledge to determine future learning. For example, having a rich and
organized knowledge about the domain of dinosaurs would facilitate the
acquisition of new dinosaur-related words (Chi \& Koeske, 1983). In
contrast, on the EXT approach, learning is driven by the connectivity of
words that are not known yet (Figure 1, right). Thus, the relevant
network structure is not internally represented by children, and the
observed connectivity effect might be an epiphenomenon of some
properties of the linguistic input. For example, highly connected words
in the input could be more easily learned because of their contextual
diversity (e.g., Smith \& Yu, 2008) or because caregivers emphasize such
words in child-directed speech (MacWhinney, 2014).

@@

Hills et al. (2009) investigated the growth of lexico-semantic networks
in toddlers and found that growth did not proceed according to INT as
was originally hypothesised by Steyvers and Tenenbaum (2005), but rather
according to
EXT.\footnote{Besides INT and EXT, the authors tested a third mechanism (called the lure of associates) which resembles EXT in that it is driven by the connectivity of external nodes, except that this connectivity is computed with respect to words that are known. However, EXT is the externally-driven scenario that best predicted the data in this previous work.}
This is a profound finding because it suggests that learning in the
early stages is mostly driven by properties of the external input,
regardless of how past knowledge is organized. However, this work has
been done in the special case of networks that are based on 1) semantic
associations, 2) production-based vocabularies, and 3) data from
English-learning children, only. The extent to which this result depends
on the domain (e.g., semantic vs.~phonological connectivity), the
vocabulary measure (production vs.~comprehension) and culture/language
is still unclear. In this work, we test the generality of the finding
along these three dimensions.

@@

We test the growth scenarios using parent reports of children
vocabularies and their normative age of acquisition (Fenson et al.,
1994). Children may vary in their individual learning trajectories, but
the aggregate data leads to an average learning pattern which is highly
consistent. It is this normative trajectory that we model in the current
study, following in this regard the steps of previous research using
similar datasets (Braginsky et al., 2016; J. C. Goodman et al., 2008;
Hills et al., 2010, 2009; Stella et al., 2017; Storkel, 2009)

@@

We used data from Wordbank (Frank, Braginsky, Yurovsky, \& Marchman,
2017), an open repository aggregating cross-linguistic language
developmental data of the MacArthur-Bates Communicative Development
Inventory (CDI), a parent report vocabulary checklist. Parent report is
a reliable and valid measure of children's vocabulary that allows for
the cost-effective collection of datasets large enough to test
network-based models of acquisition (Fenson et al., 1994). When filling
out a CDI form, caregivers are either invited to indicate whether their
child \enquote{understands} (comprehension) or \enquote{understands and
says} (production) each of about 400-700 words. For younger children
(e.g., 8 to 18 months in the English data), both comprehension and
production are queried, whereas for older children (16 to 36 months)
only production is queried. Due to these limitations, we use data from
younger children to test comprehension and data from older children to
test production. Following previous studies (Hills et al., 2009;
Storkel, 2009), we restrict our analysis to the category of
nouns.\footnote{This choice was also forced by the fact that the great majority of association data produced by adults–and which we used to build semantic networks–were nouns.}.
Table \ref{tab:stats} gives an overview of the data we used.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{ms_files/figure-latex/corrAll-1} \caption{Age of production (A) and comprehension (B) in the global network as predicted by the degree (i.e., connectivity) in this network. Results are shown in each language for phonological and semantic networks. Each point is a word, with lines indicating linear model fits, and numbers indicating the Pearson correlation coefficients.}\label{fig:corrAll}
\end{figure}

@@

Figure \ref{fig:corrAll} shows how the age of production and
comprehension for each word varies as a function of its degree (or
indegree for the semantic networks) as well as the correlation values.
For ease of visual comparison, the predictor (i.e., the degree) was
centered and scaled across languages. The plots show, overall, a
negative correlation between the month of acquisition and the degree. In
production data, the average correlation across languages was -0.24
(SD=0.10) for the semantic networks and -0.30 (SD=0.05) for the
phonological networks. In comprehension data, the average correlation
was -0.21 (SD=0.08) for the semantic networks and -0.21 (SD=0.07) for
the phonological networks. These results indicate that nouns with higher
degrees are generally learned earlier, thus replicating previous
findings in English (e.g., Storkel 2004, 2009; Hills et al. 2009) and
extending these findings to 10 different languages, generally, in both
production- and comprehension-based vocabularies.

@@

First, the results replicate Hills et al.'s original finding regarding
the semantic network in English and the production-based vocabulary,
which is that this network grows by EXT, not by INT. Second, our results
show that, generally speaking, this finding generalizes to
comprehension-based vocabulary, and holds across languages. This
generalization was obtained in both the semantic\footnote{One could
  imagine that the fact of using English free association norms
  cross-linguistically would decrease the effect of non-English semantic
  networks because of possible cultural differences. However, our
  findings do not support this assumption, rather it supports our
  initial approximation about the universality of the semantic
  similarity measure. That said, this apprimation is not perfect; there
  is evidence that a small part of the variance in free assication data
  can be explained by phonological similarity (Kachergis, Cox, \& Jones,
  2011; Matusevych \& Stevenson, 2018), thus leading to possibly minor
  cross-linguistic differences.} and phonological domains.

@@

Above we showed that the way semantic and phonological information is
structured in the learning environment contributes to noun learning (via
EXT) across languages. However, we know that other factors influence
learning as well (e.g., Braginsky et al., 2016). Next we investigated
how semantic and phonological connectivity interact with two other
factors. The first one is word frequency, a well studied factor shown to
predict the age of acquisition in a reliable fashion (e.g. J. C. Goodman
et al., 2008). The second factor is word length, which was shown to
correlate with phonological connectivity: Shorter words are more likely
to have higher connectivity (Pisoni, Nusbaum, Luce, \& Slowiaczek, 1985;
Vitevitch \& Rodríguez, 2005).

@@

For word length, we counted the number of phonemes in our generated IPA
transcription. For word frequency, we used the frequency estimates from
Braginsky et al. (2016) where unigram counts were derived based on
CHILDES corpora in each language (MacWhinney, 2014). For each word,
counts included words that shared the same stem (e.g., \enquote{cats}
counts as \enquote{cat}), or words that were synonymous (e.g.
\enquote{father} counts as \enquote{daddy}). Although these frequency
counts use transcripts from independent sets of children, they are based
on large samples, and this allows us to average out possible differences
between children and the specificities of their input (see J. C. Goodman
et al., 2008).

@@

\hypertarget{refs}{}
\hypertarget{ref-arbesman2010}{}
Arbesman, S., Strogatz, S. H., \& Vitevitch, M. S. (2010). The structure
of phonological networks across multiple languages. \emph{International
Journal of Bifurcation and Chaos}, \emph{20}(03), 679--685.

@@

@@

@@

@@

>>>>

What factors shape vocabulary learning over the course of early
childhood? To investigate this question, scientists have adopted
multiple research strategies, from conducting controlled laboratory
experiments (e.g. Markman, 1990) to analyzing dense corpora capturing
language learning in context (e.g., B. C. Roy, Frank, DeCamp, Miller, \&
Roy, 2015). One prominent strategy consists in documenting the timeline
of words' acquisition, and studying the properties that make words easy
or hard to learn (e.g., J. C. Goodman, Dale, \& Li, 2008; Huttenlocher,
Haight, Bryk, Seltzer, \& Lyons, 1991). For example, J. C. Goodman et
al. (2008) found that, within a lexical category (e.g., nouns), higher
parental frequency is associated with earlier learning. Researchers have
studied the role of several other factors such as word length and the
mean length of utterances in which the word occurs (e.g., Braginsky,
Yurovsky, Marchman, \& Frank, 2016; Swingley \& Humphrey, 2018).

Besides word-level properties, the structure of the lexicon (that is,
how words relate to one another) was also found to predict the Age of
Acquisition (AoA) of words. The lexical structure can be characterized
in terms of a network where each node represents a word in the
vocabulary, and each link between two nodes represents a relationship
between the corresponding pair of words (e.g., Collins \& Loftus, 1975;
Luce \& Pisoni, 1998). Previous studies have investigated early
vocabulary structure by constructing networks using a variety of
word-word relations including shared semantic features (McRae, Cree,
Seidenberg, \& McNorgan, 2005), target-cue relationships in free
association norms (Nelson, McEvoy, \& Schreiber, 1998), co-occurrence in
child directed speech (MacWhinney, 2014), and phonological relatedness
(Vitevitch, 2008). These studies have generally found that children tend
to produce words that have higher neighborhood density (i.e., high
connectivity in the network) earlier, both at the phonological and the
semantic level (Carlson, Sonderegger, \& Bane, 2014; Hills, Maouene,
Riordan, \& Smith, 2010; Hills, Maouene, Maouene, Sheya, \& Smith, 2009;
Stella, Beckage, \& Brede, 2017; Storkel, 2009).

@@

These two proposals represent two divergent ideas about the role of
lexical networks in acquisition. On the INT proposal, learning is driven
by known words with high connectivity to other known words (Figure 1,
left). Thus, the network structure is a causal factor in early word
learning, that is, children rely on the organization of their past
knowledge to determine future learning (Altvater-Mackensen \& Mani,
2013; Borovsky, Ellis, Evans, \& Elman, 2016; Chi \& Koeske, 1983;
Storkel, 2009). In contrast, on the EXT approach, learning is driven by
the connectivity of words that are not known yet (Figure 1, right).
Thus, the relevant network structure is not internally represented by
children, and the observed connectivity effect might be an epiphenomenon
of some properties of the linguistic input. For example, highly
connected words in the input could be more easily learned because of
their contextual diversity, allowing for an easier meaning
disambiguation (McMurray, Horst, \& Samuelson, 2012; Smith \& Yu, 2008;
Yurovsky \& Frank, 2015). Another reason could be that these words are
emphasized by the cargivers in their child-directed speech (Clark, 2007;
Hoff \& Naigles, 2002; Huttenlocher et al., 1991).

@@

Hills et al. (2009) investigated the growth of lexico-semantic networks
in toddlers and found that growth did not proceed according to INT as
was originally hypothesised by Steyvers and Tenenbaum (2005), but rather
according to
EXT.\footnote{Besides INT and EXT, the authors tested a third mechanism (called the lure of associates) which resembles EXT in that it is driven by the connectivity of external nodes, except that this connectivity is computed with respect to words that are known. However, EXT is the externally-driven scenario that best predicted the data in this previous work.}
This is an important finding because it suggests that learning in the
early stages is mostly driven by properties of the external input,
regardless of how past knowledge is organized. However, this work has
been done in the special case of networks that are based on 1) semantic
associations, 2) production-based vocabularies, and 3) data from
English-learning children, only. The extent to which this result depends
on the domain (e.g., semantic vs.~phonological connectivity), the
vocabulary measure (production vs.~comprehension) and culture/language
is still unclear. In this work, we test the generality of the finding
along these three dimensions.

@@

We adopted the following research strategy. We used a cross-linguistic
dataset of children's vocabulary development (Fenson et al., 1994;
Frank, Braginsky, Yurovsky, \& Marchman, 2017), and we studied the
timeline of word learning using the normative/average age of acquisition
(i.e., the age at which at least 50\% of children know a given words).
Our choice of studying the normative learning trajectory rather than the
individual trajectories was motivated by the nature of the dataset
used---which is primarily based on cross-sectional studies. Children may
vary in their individual learning trajectories, but the aggregate data
provide highly robust measures of the average learning patterns (Fenson
et al., 1994). The use of such measures has lead to important insights
on the mechanisms of word learning (Braginsky et al., 2016; J. C.
Goodman et al., 2008; Hills et al., 2010, 2009; Stella et al., 2017;
Storkel, 2009).

@@

We used data from Wordbank (Frank et al., 2017), an open repository
aggregating cross-linguistic language developmental data of the
MacArthur-Bates Communicative Development Inventory (CDI), a parent
report vocabulary checklist. Parent report is a reliable and valid
measure of children's vocabulary that allows for the cost-effective
collection of datasets large enough to test network-based models of
acquisition (Fenson et al., 1994). When filling out a CDI form,
caregivers are either invited to indicate whether their child
\enquote{understands} (comprehension) or \enquote{understands and says}
(production) each of about 400-700 words. For younger children (e.g., 8
to 18 months in the English data), both comprehension and production are
queried, whereas for older children (16 to 36 months) only production is
queried. Due to these limitations, we use data from younger children to
test comprehension and data from older children to test production.
Following previous studies (Hills et al., 2009; Storkel, 2009), we
restrict our analysis to the category of
nouns.\footnote{This choice was also forced by the fact that the great majority of association data produced by adults–and which we used to build semantic networks–were nouns.}.
Table \ref{tab:stats} gives an overview of the data we used.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{ms_files/figure-latex/corrProd-1} \caption{Age of production in the global network as predicted by the degree (i.e., connectivity) in this network. Results are shown in each language for phonological and semantic networks. Each point is a word, with lines indicating linear model fits, and numbers indicating the Pearson correlation coefficients.}\label{fig:corrProd}
\end{figure}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{ms_files/figure-latex/corrComp-1} \caption{Age of comprehension in the global network as predicted by the degree (i.e., connectivity) in this network. Results are shown in each language for phonological and semantic networks. Each point is a word, with lines indicating linear model fits, and numbers indicating the Pearson correlation coefficients.}\label{fig:corrComp}
\end{figure}

@@

Figures \ref{fig:corrProd} and \ref{fig:corrComp} show how the age of
production and comprehension, respectively, correlates with the degree
(or indegree for the semantic networks). For ease of visual comparison,
the predictor (i.e., the degree) was centered and scaled across
languages. The plots show, overall, a negative correlation between the
month of acquisition and the degree. In production data, the average
correlation across languages was -0.24 (SD=0.10) for the semantic
networks and -0.30 (SD=0.05) for the phonological networks. In
comprehension data, the average correlation was -0.21 (SD=0.08) for the
semantic networks and -0.21 (SD=0.07) for the phonological networks.
These results indicate that nouns with higher degrees are generally
learned earlier, thus replicating previous findings in English (e.g.,
Storkel 2004, 2009; Hills et al. 2009) and extending these findings to
10 different languages, generally, in both production- and
comprehension-based vocabularies.

@@

First, the results replicate Hills et al.'s original finding regarding
the semantic network in English and the production-based vocabulary,
which is that this network grows by EXT, not by INT. Second, our results
show that, generally speaking, this finding generalizes to
comprehension-based vocabulary, and holds across languages. This
generalization was obtained in both the semantic\footnote{One could
  imagine that the fact of using English free association norms
  cross-linguistically would decrease the effect of non-English semantic
  networks because of possible cultural differences. However, our
  findings do not support this assumption, rather it supports our
  initial approximation about the universality of the semantic
  similarity measure. That said, this apprimation is not perfect; for
  example there is evidence that a small part of the variance in free
  assication data can be explained by phonological similarity
  (Kachergis, Cox, \& Jones, 2011; Matusevych \& Stevenson, 2018), thus
  leading to possibly minor cross-linguistic differences.} and
phonological domains.

@@

Above we showed that the way semantic and phonological information is
structured in the learning environment contributes to noun learning (via
EXT) across languages. However, we know that other factors influence
learning as well (e.g., Braginsky et al., 2016). Next we investigated
how semantic and phonological connectivity interact with two other
factors. The first one is word frequency, a well studied factor shown to
predict the age of acquisition in a reliable fashion (e.g. J. C. Goodman
et al., 2008). The second factor is word length, which was shown to
correlate with phonological connectivity: Shorter words are more likely
to have higher connectivity (D. B. Pisoni, Nusbaum, Luce, \& Slowiaczek,
1985; Vitevitch \& Rodríguez, 2005).

@@

For word length, we counted the number of phonemes in our generated IPA
transcription. For word frequency, we used the frequency estimates from
Braginsky et al. (2016) where unigram counts were derived based on
CHILDES corpora in each language (MacWhinney, 2014). For each word,
counts included words that shared the same stem (e.g., \enquote{cats}
counts as \enquote{cat}), or words that were synonymous (e.g.
\enquote{father} counts as \enquote{daddy}). Although these frequency
counts use transcripts from independent sets of children, they are based
on large samples, and this allows us to average out possible differences
between children and the specificities of their input (see J. C. Goodman
et al., 2008 for a similar research strategy).

@@

\hypertarget{refs}{}
\hypertarget{ref-altvater2013}{}
Altvater-Mackensen, N., \& Mani, N. (2013). Word-form familiarity
bootstraps infant speech segmentation. \emph{Developmental Science},
\emph{16}(6).

\hypertarget{ref-arbesman2010}{}
Arbesman, S., Strogatz, S. H., \& Vitevitch, M. S. (2010). The structure
of phonological networks across multiple languages. \emph{International
Journal of Bifurcation and Chaos}, \emph{20}(03), 679--685.

@@

\hypertarget{ref-borovsky2016}{}
Borovsky, A., Ellis, E. M., Evans, J. L., \& Elman, J. L. (2016).
Lexical leverage: Category knowledge boosts real-time novel word
recognition in 2-year-olds. \emph{Developmental Science}, \emph{19}(6).

@@

\hypertarget{ref-huttenlocher1991}{}
Huttenlocher, J., Haight, W., Bryk, A., Seltzer, M., \& Lyons, T.
(1991). Early vocabulary growth: Relation to language input and gender.
\emph{Developmental Psychology}, \emph{27}(2).

@@

\hypertarget{ref-luce1998}{}
Luce, P. A., \& Pisoni, D. B. (1998). Recognizing spoken words: The
neighborhood activation model. \emph{Ear and Hearing}, \emph{19}(1).

@@

\hypertarget{ref-mcmurray2012}{}
McMurray, B., Horst, J. S., \& Samuelson, L. K. (2012). Word learning
emerges from the interaction of online referent selection and slow
associative learning. \emph{Psychological Review}, \emph{119}.
